---
title: "Simple NN"
author: "Lukasz A. Bartnik"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple NN}
  %\VignetteEngine{knitr::bookdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(collapse = TRUE, comment = NA, prompt = FALSE, echo = TRUE, cache = TRUE)
```

```{r}
library(dnnteach)
library(ggplot2)
library(tibble)
```



What are the mechanics of optimizing a NN? Let's take a closer look at a very simple NN - in fact, so simple that some people wouldn't even call it a NN - and use it to gain a few insights into what exactly happens during a gradient-descent-based optimization.

Our NN will attempt to follow the outputs of a simple quadratic function (we will, thus, have a single input `x`):

$$f(x) = x^2$$ 
```{r}
tibble(x = seq(-5, 5, .1), `f(x)` = quadratic(x)) %>%
  ggplot() + geom_line(aes(x = x, y = `f(x)`))
```


We will start with a two-neuron NN. Since it is supposed to approximate $f(x)$ we will denote it as $\hat{f}(x)$:

$$
\begin{equation}
\hat{f}(x) = \sum_{i=1}^{2} \sigma(w_i x + b_i)
(\#eq:quadratic)
\end{equation}
$$
where $\sigma(x)$ is the sigmoid function:

$$\sigma(x) = \frac{1}{1 + e^{-x}}$$
I said that our NN is unusually simple - and it is, as we sum the sigmoids without any weights. Thus, this NN has only four parameters: $(w_1, w_2, b_1, b_2)$!


The sigmoid function looks like this:

```{r}
tibble(x = seq(-6, 6, .1), y = sigma(x)) %>%
  ggplot() + geom_line(aes(x = x, y = y))
```

Since we have two of those, my first guess is that the ideal NN would take one as is and the other one reversed along the $x$ axis - like so:

```{r}
tibble(x = seq(-6, 6, .1), y1 = sigma(x-2), y2 = sigma(-x-2)) %>%
  ggplot() +
  geom_line(aes(x = x, y = y1)) +
  geom_line(aes(x = x, y = y2), color = 'red')
```

Once summed up according to Equation \@ref(eq:quadratic), we will get something that resembles a quadratic function, at least in the range $(-3, 3)$
(you might notice that this an equivalent of the following weight values: $w_1=1$, $w_2=-1$, $b_1=b_2=-2$):

```{r}
tibble(x = seq(-6, 6, .1), y = sigma(x-2) + sigma(-x-2)) %>%
  ggplot() + geom_line(aes(x = x, y = y))
```


So, what does it take for NN to arrive at this solution?



```{r}
X <- runif(100, -2, 2)
Y <- quadratic(X) + rnorm(length(X), sd=.1)
```

```


if (FALSE) {
  experiment(X, Y, epochs = 500, w = c(-2, 2))

  experiment(X, Y, epochs = 1000, D = 2, eta = 0.01)
}



if (FALSE) {
  ANS <- grid(X, Y, c(-1, -1))

  hist(log(ANS$Loss))

  pdf("by_Loss.pdf")
  ANS %>%
    mutate(ll = log(Loss)) %>%
    arrange(ll) %>%
    a_ply(1, function(ans) {
      print(result2(X, Y, ans) + ggtitle(ans$ll))
    })
  dev.off()

  ANS %>%
    mutate(ll = log(Loss)) %>%
    ggplot() +
    geom_point(aes(x = w1, y = w2, color = ll))

  ANS %>%
    mutate(ll = log(Loss)) %>%
    ggplot() +
    geom_point(aes(x = b1, y = b2, color = ll))

  classify100 <- function(w1, w2) {
    if (w1 > 3 && w2 < -3) return(1)
    if (w1 < -3 && w2 > 3) return(1)
    if (w1 < -3 && w2 < 1) return(2)
    if (w1 < 1 && w2 < -2) return(2)
    if (w1 > 2 && abs(w2) < .5) return(3)
    if (abs(w1) < .5 && w2 > 2) return(3)
    if (w1 > 2 && w2 > 2) return(4)
    return(5)
  }

  classify250 <- function(w1, w2) {
    if (w1 > 3 && w2 < -3) return(1)
    if (w1 < -3 && w2 > 3) return(1)
    if (w1 < -2 && w2 < -2) return(2)
    if (w1 < -3 && w2 < 1) return(3)
    if (w1 < 1 && w2 < -3) return(3)
    if (w1 > 2 && w2 > 2) return(4)
    #    if (w1 > 3 && abs(w2) < .5) return(4)
    #    if (abs(w1) < .5 && w2 > 3) return(4)
    return(6)
  }


  ANS2 <- ANS %>%
    ddply(.(no), function(run) {
      run$group <- classify250(tail(run, 1)$w1, tail(run, 1)$w2)
      run
    }) %>%
    mutate(group = as.factor(group))


  ANS2 %>%
    filter(stage == 'finish') %>%
    ggplot() +
    geom_point(aes(x = w1, y = w2, color = group))

  ANS2 %>%
    as_tibble %>%
    filter(i %% 10 == 0 | stage %in% c("finish", "start")) %>%
    ggplot() +
    geom_point(aes(x = w1, y = w2, color = group, size=stage), alpha = .1) +
    scale_size_manual(values=c(1.3, .05, 1.3)) +
    scale_shape_manual(values = c(21, 16, 2))

  ggsave("trace_250_dense.pdf", width = 12, height = 10)


}
```
